ISODS: An ISODATA-Based Initial Centroid Algorithm

### 2021-01-08 paper/paper.tex
adjust algorithm details, finish proposed method section, start results

\pretitle{\begin{center}\Huge\bfseries} % Article title formatting
\posttitle{\end{center}} % Article title closing formatting
\title{ISO-S: An ISODATA-Based Initial Centroid Algorithm} % Article title
\author{%
  \textsc{
    Noah Gardner$^1$\thanks{ngardn10@students.kennesaw.edu}, 
    Sahar Yarmohamadi$^2$\thanks{yarmohamadishr@gmail.com},
    Anthony Phan$^1$\thanks{aphan5@students.kennesaw.edu},
    Chih-Cheng Hung$^1$\thanks{chung1@students.kennesaw.edu}}\\[1ex]
  \normalsize $^1$Kennesaw State University \\
  \normalsize College of Computing and Software Engineering \\
  \normalsize Marietta, GA, USA
  \and
  \normalsize $^2$Institution 2 \\ %TODO 
}
\date{} % Leave empty to omit a date

@@

\section{Methods}

@@

K-Means algorithm is an unsupervised clustering algorithm and aims to partition
$N$ observations into $K$ clusters. Data points in the same cluster are very similar
to and with less variation. The user defines $K$, and these $K$ centers are placed
as far as possible for obtaining better clustering results. K-Means algorithm is
susceptible to initial starting cluster centers, and different initial values
result in a different classification. The next step is to assign each data point
to the nearest cluster center using Euclidean-based distance or other measures
based on the application-specific. After the assignment, the cluster centers are
updated by taking the average of all data points that belong to each cluster. We
continue this process until we meet the convergence criteria or the last
iteration. The algorithm is as follow:

\begin{itemize}
\item \textbf{Step 1:} Choose $K$ cluster centers from the data set.
\item \textbf{Step 2:} Assign each element of the data set to the cluster based
on the minimum distance from the cluster center.
\item \textbf{Step 3:} Re-calculate new cluster centers for each cluster after
the assignment in step 2, e.g., computing the mean of all patterns assigned to
each cluster.
\item \textbf{Step 4:} Continue steps 2 and 3 until Clustering convergence
occurs, or the number of iterations is exceeded.
\end{itemize}

For selecting the initial cluster centers in \textbf{Step 1}, there are two
standard methods:

\begin{itemize}
\item \textbf{Random Centers:} Explain here %TODO
\item \textbf{K-Means++:} Explain here %TODO
\end{itemize}

@@

ISODATA is another frequently used clustering algorithm and is similar to the
K-Means algorithm. However, in the ISODATA algorithm, the number of cluster
centers is determined dynamically, and it allows for a different number of
cluster centers. In contrast, the K-Means algorithm assumes that the number of
cluster centers is specified and fixated by the user. Some parameters such as
initial clusters' means, the minimum number of patterns in each cluster,
splitting and lumping parameters, and the number of iterations should be defined
and specified by the user. Like the K-Means algorithm, the first steps are
assignment all data points to the nearest cluster center and then recalculating
the cluster centers by replacing the average data point in each cluster. In the
next step, if a cluster does not reach the minimum number of clusters, the
user-specified, we eliminate it. After that, we do the splitting or lumping
process if their conditions are met. The cluster centers are updated based on
the newly grouped patterns. We continue this process until the convergence
occurs or the number of iterations is exceeded. The ISODATA algorithm is as
follow:

\begin{itemize}
\item \textbf{Step 1:} Choose $K$ cluster centers from the data set.
\item \textbf{Step 2:} Assign each element of the data set to the cluster based
on the minimum distance from the cluster center.
\item \textbf{Step 3:} Re-calculate new cluster centers for each cluster after
the assignment in step 2, e.g., computing the mean of all patterns assigned to
each cluster.
\item \textbf{Step 4:} Compare the number of patterns in each cluster against
the user's number ($\theta n$)and discard small sets. 
\item \textbf{Step 5:} Do splitting process based on splitting parameter
($\theta e$) if this iteration is an odd-numbered iteration of the number of
clusters is less than or equal to ½ the desired.
\item \textbf{Step 6:} Do merging process based on merging parameter ($\theta
c$)if this iteration is even-numbered, or if the number of clusters is greater
than or equal to twice the number desired.
\item \textbf{Step 7:} Start the process again at step 2 with new cluster
centers after splitting or lumping.
\item \textbf{Step 8:} If this is not the last iteration, start the process
again at step 2. If this is the last iteration, then end the process.
\end{itemize}

@@

\subsection{Proposed Method: ISO-S}
The proposed clustering algorithm's objective is to improve the result we obtain
from the ISODATA algorithm by decreasing the number of iterations needed for
convergence. Inspired by the ISODATA algorithm, this capability is achieved
using the splitting technique to update each iteration cluster centers. The main
inputs for this algorithm are the arbitrary initial number of cluster centers
and the processing parameters. First, the algorithm starts with the one cluster
center and assign the patterns to this cluster center. Second, the new algorithm
updates the cluster center by computing the mean of data points. Finally, the
algorithm employs a splitting method to form new cluster centers. The cluster
centers with the smallest number of members are discarded at the end of the
maximum number of cluster centers we desire to obtain an odd number. We continue
this process until we meet the convergence criteria: the maximum number of
clusters desired or until we reach the maximum number of iterations. The
algorithm is outlined below:

@@

>>>>

\pretitle{\begin{center}\Huge\bfseries} % Article title formatting
\posttitle{\end{center}} % Article title closing formatting
\title{ISODS: An ISODATA-Based Initial Centroid Algorithm} % Article title
\author{%
  \textsc{
    Noah Gardner$^1$\thanks{ngardn10@students.kennesaw.edu}, 
    Sahar Yarmohammadtoosky$^2$\thanks{yarmohamadishr@gmail.com},
    }\\[1ex]
  \textsc{
    Anthony Phan$^1$\thanks{aphan5@students.kennesaw.edu},
    Chih-Cheng Hung$^1$\thanks{chung1@students.kennesaw.edu}} \\[1ex]
  \normalsize $^1$Kennesaw State University \\
  \normalsize College of Computing and Software Engineering \\
  \normalsize Marietta, GA, USA \\[1ex]
  \normalsize $^2$Iran University of Science and Technology \\
  \normalsize Tehran, Iran
}
\date{} % Leave empty to omit a date

@@

@@

The K-Means algorithm is an unsupervised clustering algorithm and aims to
partition $N$ observations into $K$ clusters [CITE]. Data points in the same
cluster are very similar to each other and have less variation compared to data
points assigned to other clusters. The user defines $K$, the number of expected
clusters. Following is a description of the K-Means algorithm:

\begin{itemize}
\item \textbf{Step 1:} Choose $K$ cluster centers from the data set.
\item \textbf{Step 2:} Assign each element of the data set to the cluster based
on the minimum distance from the cluster center.
\item \textbf{Step 3:} Re-calculate new cluster centers for each cluster after
the assignment in \textbf{Step 2} by computing the mean of all patterns
assigned to each cluster.
\item \textbf{Step 4:} Repeat \textbf{Step 2 and Step 3} until clustering
convergence occurs, or the set number of iterations is exceeded.
\end{itemize}

For selecting the initial cluster centers in \textbf{Step 1}, there are two
standard methods: Random Centers, and K-Means++ [CITE]. In Random Centers, $K$
samples are randomly chosen from the dataset to serve as the initial cluster
centers. In K-Means++, $1$ sample is randomly chosen from the dataset to serve
as the initial cluster center. Then, $K-1$ samples are iteratively selected,
where the probability of a sample being chosen as a cluster center is the
distance of the sample from previously chosen centers [CITE]. The K-Means
algorithm is susceptible to initial starting cluster centers, where different
initial values can result in a different classification.

@@

ISODATA is another frequently used clustering algorithm and is based on the
K-Means algorithm. Like K-Means, cluster centers are iteratively updated and
reassigned to the samples in the dataset. However, ISODATA determines the number
of clusters dynamically and allows for a different number of cluster centers. In
contrast, the K-Means algorithm requires a fixed user-set parameter of the
number of expected clusters. The advantage found within ISODATA allows for
discovery of a different number of clusters, which is achieved through
discarding clusters with a small number of member samples, and merging and
splitting clusters that meet certain parameters. There are three main ISODATA
parameters:

\begin{itemize}
\item $\theta n$ - minimum number of cluster members, which determines whether
the cluster is discarded,
\item $\theta e$ - maximum standard deviation allowed for a cluster, which
determines whether the cluster is split,
\item $\theta c$ - minimum distance between clusters, which determines whether
two clusters are merged.
\end{itemize}

Additionally, the user must set $K$, the maximum number of clusters to generate.
The ISODATA algorithm is briefly explained below:

\begin{itemize}
\item \textbf{Step 1:} Choose $K$ cluster centers from the data set.
\item \textbf{Step 2:} Assign each element of the dataset to the cluster based
on the minimum distance from the cluster center.
\item \textbf{Step 3:} Re-calculate new cluster centers for each cluster by
computing the mean of all patterns assigned to each cluster.
\item \textbf{Step 4:} Discard clusters that have fewer than ($\theta n$)
members. If any clusters were discarded, move to \textbf{Step 7}.
\item \textbf{Step 5:} If this iteration is an odd numbered iteration or the
number of clusters is less than or equal to  $1/2 * K$, then split clusters
which have a standard deviation in a dimension greater than ($\theta e$). Move
to \textbf{Step 7}.
\item \textbf{Step 6:} If this iteration is an even numbered iteration or the
number of clusters is greater than or equal to $2 * K$, then merge clusters which
have an intercluster distance less than ($\theta c$).
\item \textbf{Step 7:} If this is not the last iteration, start the process
again at step 2. If this is the last iteration or clustering convergence occurs,
then end the process.
\end{itemize}

There are some disadvantages to the ISODATA algorithm. First, ISODATA is
sensitive to the parameters $\theta n$, $\theta e$, and $\theta c$. These
parameters should be fine-tuned for optimal performance based on the attributes
of the dataset being tested. In addition, similar to K-Means, it is sensitive to
initial cluster centers, although less so than K-Means due to the extra
processes of discarding, splitting, and merging. The initial centroids can
affect both the quality of clusters produced and the number of iterations
required to converge [CITE].

@@

\section{Proposed Method}

The proposed clustering algorithm's objective is to improve the result we obtain
from the ISODATA algorithm by targeting the initial cluster centers. Inspired by
the ISODATA algorithm, Iterative Self-Organizing Data Split (ISODS) uses the
splitting process from ISODATA to create the required initial cluster centers to
use for the ISODATA algorithm. The inputs for the algorithm are the number of
cluster centers $K$ as an integer, and the data to generate centers from. The
algorithm starts by selecting $1$ sample as an initial cluster center. In the
next step, each data is assigned to the cluster and the mean of the cluster is
updated. Then, by the ISODATA splitting process, each cluster is split along the
dimension with the highest standard deviation. The assignments for the new
clusters are calculated based on distance and the centers are updated. This
process is continued until either the number of clusters is greater than or
equal to $K$. If the number of clusters is greater than $K$, the clusters with
the fewest members are discarded one at a time, and the assignments and centers
are updated after each rejection. The algorithm is outlined below:

\begin{itemize}
\item \textbf{Step 1:} Choose $1$ observation as a cluster center from the dataset.
\item \textbf{Step 2:} Assign each element of the dataset to the cluster based
on the minimum distance from the cluster center.
\item \textbf{Step 3:} Re-calculate new cluster centers for each cluster by
computing the mean of all patterns assigned to each cluster.
\item \textbf{Step 4:} If the number of clusters is equal to $K$, then the
process is terminated. If the number of clusters is greater than $K$, then the
clusters with the fewest members is discarded, and the algorithm moves to
\textbf{Step 2}.
\item \textbf{Step 5:} Calculate the standard deviation of each cluster in each
dimension. Then, split each cluster along the dimension with the greatest
standard deviation.
\end{itemize}

This method began as a modified ISODATA process. However, the modification was
unique in that there is a clear distinction from the proposed change and the
beginning of the original process. Although more complex than other methods of
generating initial cluster centers, this approach guarantees variance between
clusters. As a result, ISODS can be used to generate clusters for other
clustering algorithms, such as K-Means. In the next section, ISODS is applied to
both ISODATA and K-Means on some benchmark UCI datasets [CITE] and compared with
Random Centers and K-Means++ Centers.

@@

Due to the response of the ISODATA algorithm to the discarding, merging, and
splitting parameters, these parameters should be carefully selected. Since the
goal of this research is to explore the impact of initial cluster centers on
ISODATA, we use a relatively simple approach for generating ISODATA parameters.
Based on the available training data with included labels, each class is
summarized with the mean and standard deviation of that class, as well as the
number of members in each class. To determine $\theta n$, the minimum number of
cluster members used for discarding, we find the class with the fewest members
and use the number of members in that class as $\theta n$. For $\theta e$, the
maximum standard deviation allowed for a cluster, we follow a similar approach
and find the maximum standard deviation within a class. Finally, for $\theta c$,
we calculate the distance between the means of each class, and use the minimum
distance as $\theta c$.
